Galaxy Imaging with Generative Models: Insights from a Two-Models Framework

### 2025-03-15 MNRAS/galaxy-generative-models-overleaf-submit-2nd/article_MNRAS.tex
section 3.3.x cont.

{\color{red}The central challenge in generation lies in estimating the \textit{score}, the key ingredient of the generation process.} Two main approaches have been proposed: (1) deriving a model for $p_t$ that incorporates regularity patterns and correlations to learn the \textit{score} via \textit{score matching} \citep{hyvarinen2005a}, or (2) employing a \textit{denoising} neural network, as introduced by \cite{Bengio2013}, further developed in \cite{Sohl-Dickstein2015,Ho2020,song2020score}, and recently examined from a mathematical perspective in \cite{kadkhodaie2024generalization}. The first approach has been utilized in works such as \cite{Guth2022b,Lempereur2024} with Gibbs energy parameterization. The second has found applications in astrophysics, including galaxy image generation \citep{smith2021}, generation of 21~cm luminosity temperature maps \citep{Zhao2023}, super-resolution of large-scale cosmic structures \citep{Schanz2023}, and Bayesian posterior sampling for weak lensing mass-mapping problems \citep{Remy2023}.

@@

{\color{red}
In the present study, we focus on galaxy image generation, taking as a reference the experiment conducted by \cite{kadkhodaie2024generalization}, who, however, used widely known machine learning datasets. All their images were grayscale (single channel), with intensity values in the range $[0,1]$ and a resolution mostly of $80 \times 80$ pixels.

@@

Thus, we restrict ourselves to relatively modest image sizes compared to those produced by models such as \texttt{light-weight-gan} and \texttt{Glow}. Our primary objective is to investigate whether the transition from \textit{memorization} to \textit{generalization} in diffusion models occurs around $10^5$ training images, as mentioned earlier, and whether this transition aligns with the number of model parameters (see Section~\ref{sec-some-comments}). Furthermore, we aim to examine how GAN and flow-based models behave as a function of dataset size. However, even when using similar image sizes, it remains an open question whether the same number of training images is required to achieve the generalization transition when working with face or bedroom images compared to galaxy images. The complexity of such images can be estimated based on the compression ratio achieved using the same algorithm\footnote{For example, \texttt{JPEG} compression with default parameters in the \texttt{PIL} library v11.1.0 (\url{https://pillow.readthedocs.io/}).}. We find that the galaxy images used in this study exhibit a complexity reduction of approximately $50\%$. With less complex images, one might expect an easier task for generative models. At the very least, this question can be explored within the framework described here.
}

@@

Figure~\ref{fig-UNet-denoising-train} presents the denoising results of two networks trained on datasets of different sizes, using an original image shared by both training sets. In contrast, Figure~\ref{fig-UNet-denoising-test} demonstrates the performance on an input image that was never included in the training data of either network, referred to as a \textit{validation image}. For both figures, the peak signal-to-noise ratio (PSNR) relative to the original image is provided above each image. {\color{red} This metric is appropriate, as it directly relates to the loss function used to optimize the denoiser. Additionally, in the context of score estimation, we focus on the overall denoising performance across the entire image, not just the galaxy pixels.} The top row shows the input images (original or noisy) fed to the denoiser. The middle row displays the outputs from the \texttt{U-Net} trained with $N=100$ images, while the bottom row shows outputs from the \texttt{U-Net} trained with $N=100,000$ images. It is important to note that the networks were trained using noisy images without access to PSNR values or noise variance information.

@@

Interestingly, even for relatively simple, low-resolution galaxy images, approximately $10^5$ samples are required to reliably achieve the transition into the \textit{generation regime}. However, for practical applications, as emphasized in \cite{kadkhodaie2024generalization}, producing higher-resolution images does not demand datasets that are orders of magnitude larger than those needed for $64 \times 64$ images. This is attributed to the hierarchical and multi-scale structure inherent to the images.

@@

Following the two-models test described in Section~\ref{sec-two_models-Diffusion}, we trained two flow-based models ("A" and "B") with independent galaxy datasets of size $N=10^5$. The cosine similarity results are shown in Figure~\ref{fig-cosine-glow}. We can draw some key observations from these results:
\begin{itemize}
    \item The generated samples appear distinct from any training image (or their augmented versions) as the right histogram does not peak at a cosine similarity of 1. 
    \item Even when seeded with the same white noise, the images generated by the two models are different, as indicated by the absence of a peak at cosine similarity 1 in the left histogram.
\end{itemize}

@@

However, Figure~\ref{fig-gan-hinge-loss-epoch100-ntrain-10p3-10p4} presents the results of the "discriminator test" for models trained with datasets of sizes $N=10^4$ (panel $(a)$) and $N=10^3$ (panel $(b)$), evaluated at \textit{epoch} 100. Readers are encouraged to compare these findings with the corresponding panel in Figure~\ref{fig-gan-hinge-loss-evol}. When training with $N=10^4$ or $N=10^3$, the results indicate that the discriminators are not entirely agnostic to the origin of the image samples, as evidenced by the distinct distributions of $\ell(D_A(\bm{x}_A), D_A(\bm{x}_B))$ and \(\ell(D_A(\bm{x}_A), D_A(\bm{x}_{\text{train}}))\) (and similarly for the $D_B$ discriminator). 
Moreover, the discrepancies in these loss distributions persist even when training continues beyond \textit{epoch} 100, suggesting that the observed divergences are not merely attributable to early convergence issues. It could be argued that these inconsistencies may arise from suboptimal hyperparameter choices in designing the \texttt{light-weight-gan} architecture used for the experiment, rather than solely from the limited dataset size. Notably, in the absence of convergence across the four loss distributions, there is no \textit{a priori} method to guide further investigations or remedial measures effectively.
%
\begin{figure}
    \centering
	\includegraphics[width=0.7\linewidth]{fig-gini-m20-gan-compare_0.pdf}
	\caption{Some morphological coefficients (see Figure~\ref{fig-morpho-coeff}) computed using \texttt{light-weight-gan} models trained with different dataset sizes ($N=10^5, 10^4, 10^3$).}
	\label{fig-morpho-coeff-gan-comparison}
\end{figure}
\begin{figure}
    \centering
        \begin{subfigure}[b]{\columnwidth}
        \centering
        \includegraphics[width=0.8\linewidth]{fig-gan-loss_discri_10000_A_B_num_100.pdf}
        \caption{$N=10^4$}
        \end{subfigure}%
    \hfill
        \begin{subfigure}[b]{\columnwidth}
        \centering
        \includegraphics[width=0.8\linewidth]{fig-gan-loss_discri_1000_A_B_num_100.pdf}
        \caption{$N=10^3$}
        \end{subfigure}%
	\caption{Same legend as for Figure~\ref{fig-gan-hinge-loss-evol} but with different dataset sizes: $N=10^4$ (a) and $N=10^3$ (b).}
	\label{fig-gan-hinge-loss-epoch100-ntrain-10p3-10p4}
\end{figure}
%
\section{Conclusion}
\label{sec-conclusion}
%
In this study, we investigated the behavior of generative models applied to galaxy image generation, focusing on their performance under controlled experimental conditions. We compared three architectures: the \texttt{light-weight-gan} GAN-based model, the \texttt{Glow} flow-based model, and the diffusion model based on a \texttt{U-Net} denoiser (Section~\ref{sec-Exp-Models}). These models were trained and evaluated using identical datasets sourced from the SDSS DR7 survey (Section~\ref{sec-Dataset}).

>>>>

{\color{red}The central challenge in generation lies in estimating the \textit{score}, the key ingredient of the generation process.} Two main approaches have been proposed: (1) deriving a model for $p_t$ that incorporates regularity patterns and correlations to learn the \textit{score} via \textit{score matching} \citep{hyvarinen2005a}, or (2) employing a \textit{denoising} neural network, as introduced by \cite{Bengio2013}, further developed in \cite{Sohl-Dickstein2015,Ho2020,song2021scorebased,song2021maximum}, and recently examined from a mathematical perspective in \cite{kadkhodaie2024generalization}. The first approach has been utilized in works such as \cite{Guth2022b,Lempereur2024} with Gibbs energy parameterization. The second has found applications in astrophysics, including galaxy image generation \citep{smith2021}, generation of 21~cm luminosity temperature maps \citep{Zhao2023}, super-resolution of large-scale cosmic structures \citep{Schanz2023}, and Bayesian posterior sampling for weak lensing mass-mapping problems \citep{Remy2023}.

@@

{\color{red}
In the present study, we focus on galaxy image generation, taking as a reference the experiment conducted by \cite{kadkhodaie2024generalization}, who, however, used widely known machine learning datasets as face or bedroom images. All their images were grayscale (single channel), with intensity values in the range $[0,1]$ and a resolution mostly of $80 \times 80$ pixels.

@@

Thus, we restrict ourselves to relatively modest image sizes compared to those produced by models such as \texttt{light-weight-gan} and \texttt{Glow}. Our primary objective is to investigate whether the transition from \textit{memorization} to \textit{generalization} in diffusion models occurs around $10^5$ training images, as mentioned earlier, and whether this transition aligns with the number of model parameters (see Section~\ref{sec-some-comments}). Furthermore, we aim to examine how GAN and flow-based models behave as a function of dataset size.

However, even when using similar image sizes, it remains an open question whether the same number of training images is required to achieve the generalization transition when working with face or bedroom images, which may be more complex than galaxy images at the same resolution. The complexity of such images can be estimated based on the compression ratio achieved using the same algorithm\footnote{For example, \texttt{JPEG} compression with default parameters in the \texttt{PIL} library v11.1.0 (\url{https://pillow.readthedocs.io/}).}. We find that the galaxy images used in this study exhibit a higher compression ratio, approximately $50\%$ greater. Thus, for less complex images, one might expect generative models to have an easier task. At the very least, this question can be explored within the framework described here.
}

@@

Figure~\ref{fig-UNet-denoising-train} presents the denoising results of the networks trained on datasets of different sizes, using an original image shared by both training sets. In contrast, Figure~\ref{fig-UNet-denoising-test} demonstrates the performance on an input image that was never included in the training data of either network, referred to as a \textit{validation image}. For both figures, the peak signal-to-noise ratio (PSNR) relative to the original image is provided above each image. {\color{red} This metric is appropriate, as it directly relates to the loss function used to optimize the denoiser. Additionally, in the context of score estimation, we focus on the overall denoising performance across the entire image, not just the galaxy pixels.} The top row shows the input images (original or noisy) fed to the denoiser. The middle row displays the outputs from the \texttt{U-Net} trained with $N=100$ images, while the bottom row shows outputs from the \texttt{U-Net} trained with $N=100,000$ images. It is important to note that the networks were trained using noisy images without access to PSNR values or noise variance information.

@@

%15-mars-25 add the complexity from the dataset section
%Interestingly, even for relatively simple, low-resolution galaxy images, aapproximately $10^5$ samples are required to reliably achieve the transition into the \textit{generation regime}. 
{\color{red} 
Interestingly, and perhaps counterintuitively, even for relatively simple, low-resolution galaxy images---considered less complex than face or bedroom images, as discussed in Section~\ref{sec-Dataset}---approximately the same amount of $10^5$ samples is required to reliably achieve the transition into the \textit{generation regime}. However, for practical applications, as emphasized in \cite{kadkhodaie2024generalization}, producing higher-resolution images does not require a proportional increase in the number of samples. This is attributed to the hierarchical and multi-scale structure inherent to the images.
}

@@

Following the two-models test described in Section~\ref{sec-two_models-Diffusion}, we trained two flow-based models ("A" and "B") with independent galaxy datasets of size $N=10^5$. The cosine similarity results are shown in Figure~\ref{fig-cosine-glow}, {\color{red} which maintains the same settings\footnote{This results in a cut on the left side of the histograms, but our focus is on the region close to a cosine similarity of around 1.} as Figure~\ref{fig-cosine-diffusion}}. We can draw some key observations from these results:
\begin{itemize}
    \item The generated samples appear distinct from any training image (or their augmented versions) as the right histogram does not peak at a cosine similarity of 1. 
    \item Even when seeded with the same white noise, the images generated by the two models are different, as indicated by the absence of a peak at cosine similarity 1 in the left histogram.
\end{itemize}

@@

However, Figure~\ref{fig-gan-hinge-loss-epoch100-ntrain-10p3-10p4} presents the results of the "discriminator test" for models trained with datasets of sizes $N=10^4$ (panel $(a)$) and $N=10^3$ (panel $(b)$), evaluated at \textit{epoch} 100. Readers are encouraged to compare these findings with the corresponding panel in Figure~\ref{fig-gan-hinge-loss-evol}. When training with $N=10^4$ or $N=10^3$, the results indicate that the discriminators are not entirely agnostic to the origin of the image samples, as evidenced by the distinct distributions of $\ell(D_A(\bm{x}_A), D_A(\bm{x}_B))$ and \(\ell(D_A(\bm{x}_A), D_A(\bm{x}_{\text{train}}))\) (and similarly for the $D_B$ discriminator). 
Moreover, the discrepancies in these loss distributions persist even when training continues beyond \textit{epoch} 100, suggesting that the observed divergences are not merely attributable to early convergence issues. It could be argued that these inconsistencies may arise from suboptimal hyperparameter choices in designing the \texttt{light-weight-gan} architecture used for the experiment, rather than solely from the limited dataset size. Notably, in the absence of convergence across the four loss distributions, there is no \textit{a priori} method to guide further investigations or remedial measures effectively.
%
\begin{figure}
    \centering
	\includegraphics[width=0.7\linewidth]{fig-gini-m20-gan-compare_0.pdf}
	\caption{Some morphological coefficients (see Figure~\ref{fig-morpho-coeff}) computed using \texttt{light-weight-gan} models trained with different dataset sizes ($N=10^5, 10^4, 10^3$).}
	\label{fig-morpho-coeff-gan-comparison}
\end{figure}
\begin{figure}
    \centering
        \begin{subfigure}[b]{\columnwidth}
        \centering
        \includegraphics[width=0.8\linewidth]{fig-gan-loss_discri_10000_A_B_num_100.pdf}
        \caption{$N=10^4$}
        \end{subfigure}%
    \hfill
        \begin{subfigure}[b]{\columnwidth}
        \centering
        \includegraphics[width=0.8\linewidth]{fig-gan-loss_discri_1000_A_B_num_100.pdf}
        \caption{$N=10^3$}
        \end{subfigure}%
	\caption{Same legend as for Figure~\ref{fig-gan-hinge-loss-evol} but with different dataset sizes: $N=10^4$ (a) and $N=10^3$ (b).}
	\label{fig-gan-hinge-loss-epoch100-ntrain-10p3-10p4}
\end{figure}
%
\section{Conclusion}
\label{sec-conclusion}
%
In this study, we investigated the behavior of generative models applied to galaxy image generation, focusing on their performance under controlled experimental conditions. We compared three architectures: the \texttt{light-weight-gan} GAN-based model, the \texttt{Glow} flow-based model, and the diffusion model based on a \texttt{U-Net} denoiser (Section~\ref{sec-Exp-Models}). These models were trained and evaluated using identical datasets sourced from the SDSS {\color{red}DR9} survey (Section~\ref{sec-Dataset}).
